---
title: "ST661 Project - Dublin Marathon "
author: ""
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Work distribution

The report is a collaborative effort covering Introduction, Data Wrangling, Analysis and Conclusion. Each of our individual contribution to the questions formed is mentioned below.

### Introduction

The project focuses on Exploratory Data Analysis (EDA) of the Dublin Marathon Finishers dataset from the year 2022. Through data visualization and statistical analysis, we aim to reveal the patterns and insights into the performance and demographics of marathon participants. The dataset includes participant details such as names, ages, genders, club affiliations, and various race metrics, offering a comprehensive overview of their performance. Our exploration will investigate factors like overall time, gender, age and various club impact, thus providing valuable insights into the distribution and characteristics of Dublin Marathon participants.

### Data Wrangling 

The code below performs data wrangling on the dataset. It begins by loading the dataset and addressing missing values, filling the "Club" column with "UNKNOWN" for consistency. Uninformative columns are then removed. Rows indicating runners who did not finish (DNF) are separated into a different dataset for further investigation. Duplicated rows are identified and removed. Finally, rows with missing values are eliminated. The resulting "marathon_data" is now refined and ready for analysis, ensuring data integrity and addressing issues like missing, duplicated, and uninformative information.

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(GGally)

# Load DataSet
marathon_data <- read.csv('dublin2022marathon.csv')

# Identify columns with missing values
missing_cols <- colnames(marathon_data)[colSums(is.na(marathon_data)|
                                                  marathon_data == "") > 0]

# Fill missing value Club column with UNKNOWN value to avoid data discrepancy
marathon_data$Club <- ifelse(is.na(marathon_data$Club) | 
                               marathon_data$Club == "" | 
                               marathon_data$Club == " ", "UNKNOWN",
                             marathon_data$Club)

# Remove uninformative columns
marathon_data <- select(marathon_data, -c("Photo", "YouTube", "Share",
                                          "Chip.Time", "Chip.Position"))

# Move values with DNF (did not finish) data as a different dataset to investigate further based on age and gender on why they were not able to finish the marathon
dnf_rows <- apply(marathon_data, 1, function(row) any(row == "DNF"))
dnf_data <- marathon_data[dnf_rows, ]

# Remove DNF values from dataset
marathon_data <- marathon_data[!dnf_rows, ]

# Check for any duplicated values
columns_to_check <- c("First.Name", "Surname", "Gender", "Gender.Position",
                      "Category", "Category.Position", "Club",
                      "Overall.Position")

duplicate_rows <- duplicated(marathon_data[, columns_to_check])
duplicate_data <- marathon_data[duplicate_rows, ]

# Remove rows with missing values - only around 500 entries
missing_rows <- apply(marathon_data, 1,
                      function(row) any(is.na(row) | row == ""))
missing_data <- marathon_data[missing_rows, ]

marathon_data <- marathon_data[!missing_rows, ]
```

### Analysis

#### 1. Visualizing the correlation between half time and finish time

```{r, fig.height=3, fig.width=5, fig.cap="Scatterplot illustrating the correlation between time taken for first and second half of the marathon. Each point represents a participant's performance, and the smooth line indicates the overall trend.", message=FALSE, warning=FALSE}

marathon_data$Gender <- as.factor(marathon_data$Gender)
marathon_data <- marathon_data |>
  mutate(
    FinalInSec = period_to_seconds(hms(Gun.Time)),
    MidInSec = period_to_seconds(hms(HALFWAY))
    )

marathon_data <- marathon_data |>
  mutate(SecondHalfSec = FinalInSec - MidInSec) 

ggplot(marathon_data, 
       aes(x=MidInSec, y=SecondHalfSec)) + 
  geom_point() +
  geom_smooth()+
  scale_x_continuous(limits = c(4000, 15000)) + 
  scale_y_continuous(limits = c(4000, 18000)) + 
  xlab("Seconds for First Half of Race") + 
  ylab("Seconds for Second Half of Race")

pval<-t.test(marathon_data$MidInSec, marathon_data$SecondHalfSec, 
          paired = T)$p.value
```

While the performances in first half and second half race are significantly different and correlated, an athlete's second half performance tends to be worse than their first half (their difference tends to be positive on average).

#### 2. Visualising relationship between Club Affiliation and Overall Time

```{r, fig.width=6, fig.height=3.5, fig.cap= "Scatter plot depicting the relationship between club affiliation and finish time for the top 20 clubs based on median overall time. The points represent the median finish time, while the connecting lines highlight the distribution of overall time taken within each club."}

club_summary <- marathon_data %>%
  group_by(Club) %>%
  summarise(
    MeanOverall = mean(FinalInSec, na.rm = TRUE),
    MedianOverall = median(FinalInSec, na.rm = TRUE),
    MinOverall = min(FinalInSec, na.rm = TRUE),
    MaxOverall = max(FinalInSec, na.rm = TRUE)
  )

top_clubs <- club_summary %>%
  arrange(MedianOverall) %>%
  slice_head(n = 20) 

# Visualize relationship between Club Affiliation and Overall Time
ggplot(top_clubs, aes(x = reorder(Club, MedianOverall), y = MedianOverall)) +
  geom_point() +
  geom_segment(aes(xend = fct_reorder(Club, MedianOverall)), yend = 0, col = "lightblue") +
  labs(title = "Clubs Affiliation and Finish Time") +
  xlab("Club")+ ylab("Finish Time(seconds)")+
  coord_flip()

filtered_data <- marathon_data %>%
  filter(Club != "UNKNOWN")%>%
  select(Club, FinalInSec)

anova_result <- aov(FinalInSec ~ Club, data = filtered_data)
a<-summary(anova_result)
a1<-cor.test(club_summary$MeanOverall, club_summary$MedianOverall)
```

The scatterplot identifies top-performing clubs BRIDGE END A.C., ARMAGH A.C., and CORK TRACK CLUB, based on their lowest median finish times. However, the presence of UNKNOWN affiliations,as in data of joining any club is not present, especially within the top five positions, raises uncertainty about club associations. ANOVA results confirm a significant difference in mean finish times among clubs (p < 0.001), emphasizing the influence of club affiliation on performance. The strong positive correlation between Mean and Median Overall Positions signals consistent and stable performance patterns within each club.

#### 3. Visualizing Runners Finish time based on Gender

```{r,fig.height=3, fig.width=5, fig.cap="Boxplot illustrating distribution of Overall time taken by male and female participants in the marathon, with the x-axis representing the genders and the y-axis representing Overall time", message=FALSE, warning=FALSE}

t<-t.test(FinalInSec ~ Gender, data = marathon_data)

ggplot(marathon_data, aes(x = Gender, y = FinalInSec,fill = Gender)) +
  geom_boxplot() +
  labs(title = "Overall Time by Gender",
       x = "Gender",
       y = "Overall Time(seconds)")
```

The boxplot compares the Overall time taken by male and female participants and reveals that, on average, females took longer to finish. Median gun time for females is higher than males. Also, the male group exhibits greater variability in performance, with some exceptionally slower times. More outliers and greater variance in one gender can influence the average Gun time for that group. 
As the p_value <0.05, we can confirm there is a significant difference in the average finish time between male and female participants.

#### 4. Visualising Performance Metrics across various Age Categories

```{r, fig.width=6, fig.height=3, fig.cap="Barplot representing overall time of participants across various age groups and genders. Each bar corresponds to a specific age category, with different colors for distinct gender groups. Height of each bar illustrates average overall time for participants."}

marathon_data<- marathon_data |>
 select(1:21) |>
separate(Category,into=c('Category','Age'),sep=1)
 
marathon_data <- marathon_data |>
  mutate(
    FinalInMins = period_to_seconds(hms(Gun.Time))/60)

# Summarize data by Age_Category
summary_by_age <- marathon_data %>%
  group_by(Age) %>%
  reframe(
    Category,
    Mean_Time = mean(FinalInMins, na.rm = TRUE),
    Median_Time = median(FinalInMins, na.rm = TRUE),
    SD_Time = sd(FinalInMins, na.rm = TRUE),
    Total_Participants = n()
  )

# Visualize Performance Metrics
ggplot(summary_by_age, aes(x = factor(Age, levels = c('S','35','40','45','50','55','60','65','70','75','80')), 
                           y = Mean_Time, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Barplot of Overall Time by Age and Gender",
       x = "Age",
       y = "Overall Time") 

```

The plot highlights a observable correlation between performance and age in the marathon. Notably, participants in the 70-80 age group exhibit the longest completion times, while the 35-40 age group demonstrates the shortest. It's noteworthy that there are no female participants in the 80-year age group. Overall, there's a negative impact of age on performance, indicating a trend of decreasing performance with increasing age.


#### 5. Visualising the Stage Position Change for top 5 Performers

*General analysis of all members of the team*
```{r, fig.width=6, fig.height=3, fig.cap="The graph illustrates the stage position changes for the top five performers in the marathon. Each line with distinct colors represents an individual performer. X-axis corresponds to different race stages, while y-axis indicates the position of each performer at those stages."}
marathon_data<- marathon_data |>
  mutate(Overall.Position=as.numeric(Overall.Position))

top_three_performers <- marathon_data %>%
  arrange(Overall.Position) %>%
  head(5) %>% 
  mutate(across(starts_with(c("Overall.Position", "Stage.Position")),
                as.character)) %>%
  pivot_longer(cols = starts_with(c("Stage.Position", "Overall.Position")),
               names_to = "Stage",
               values_to = "Position") %>%
  mutate(Stage = forcats::fct_inorder(Stage),
         Position = as.numeric(Position))  

ggplot(top_three_performers, aes(x = Stage, y = Position, 
                                color = First.Name, 
                                group = paste(First.Name, Surname))) +
  geom_line() +
  geom_point() +
  labs(x = "Race Stage",
       y = "Position",
       title = "Stage Position Change for top 5 Performers") +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange")) +
  scale_x_discrete(labels = c("10K", "20K", "HALFWAY", "30K", "40K", "FINAL")) +
  scale_y_continuous(breaks =seq(1,12,1))

```

The plot reveals how the top 5 performers' stage positions evolve throughout the marathon. It emphasizes that the initial positions of participants don't necessarily determine their final standings. Notable examples include "Taoufik Allam," starting in 5th place but finishing 1st through consistent improvement, and "Craig Curley," who, despite an initial 12th position, secured 5th place with steady performance. This underscores the significance of sustained excellence over the race, indicating that ongoing performance matters more than the starting position alone.

### Conclusion

The analysis points to Clubs BRIDGE END A.C., ARMAGH A.C., and CORK TRACK CLUB as the most favorable choices for marathon participation, with runners in it achieving better Overall Time. Notably, a positive correlation exists between performances in the first and second halves of the race, emphasizing the significance of pacing. Male participants generally outperform their female counterparts, indicating a gender-based performance gap. Additionally, a noteworthy relationship between age and performance suggests the need for tailored training approaches for different age groups.

